The function-sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-27 crelle-dictionary.tex
Finished proof reading Section 3.

By any measure, there are more quasicharacter sheaves for $T$ than quasicharacters of $T(K)$.  
In this regard, we are reminded of the work of David Vogan \cite{vogan:93a}, in which he finds a geometrization of complete Langlands parameters for $p$-adic groups 
and, in the process, is led to study the representations of all the pure rational forms of the $p$-adic group simultaneously.
A similar phenomenon appears in recent work by Joseph Bernstein in which his geometric Ansatz leads to the study of certain sheaves on the stacky classifying space of the $p$-adic group, resulting in a category which appears to be tied to the representations of all the pure rational forms of the $p$-adic group \cite{bernstein:vogan_conference}.
Indeed, Bernstein has suggested to us that our category of quasicharacters for $T$ may be tied to quasicharacters of all the pure rational forms of $T$.

@@

\begin{proof}
Let $f$, $A$, $V$ and $\psi$ be as above; set $\qcs{E} = (f_! V_H)_\psi$.
Since $A$ is abelian, $\qcs{E}$ is an $\ell$-adic local system on $G$ of rank one.
We must show that comes equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E}\boxtimes\qcs{E}$.
To do this we use \'etale descent to see that pullback along $f$ gives an equivalence between $\ell$-adic local systems on $G$ and $A$-equivariant local systems on $H$; {\it cf.\ } \cite{bernstein-luntz:equivariant_sheaves}*{Prop 8.1.1}. 
In particular, $f^*\qcs{E}$ is the $A$-equivariant constant sheaf $V$ on $H$ with character $\psi$.
Since $f$ is a morphism of group schemes, the functor $f^*$ defines $\mu_\qcs{E} : m^*\qcs{E} \to \qcs{E}\boxtimes\qcs{E}$
from the isomorphism $m^*\psi \iso \psi \boxtimes\psi$ determined by $V\to V\otimes V$.
\end{proof}

@@

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object of $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function-sheaf dictionary for connected commutative algebraic groups over finite fields, as in \cite{deligne:SGA4.5}*{Sommes trig.} or \cite{laumon:87a}*{1.1.3}, gives us all we need, as we briefly recall. %\todo{Fix citelist} %\citelist{\cite{deligne:SGA4.5}*{Sommes trig.},\cite{laumon:87a}*{1.1.3}}
The key observation is that, by \'etale descent, pullback by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every $\qcs{E}$ in $\bQC(G)$. 
(Note that this is a particular case of the argument in the proof of Proposition~\ref{prop:finite}.)
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_! V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ of the constant sheaf $V$ on $G$ equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$. 
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
\end{proof}

%\begin{remark}
We have just seen that, for a connected commutative algebraic group $G$ over $\Fq$, the category of quasicharacter sheaves on $G$ is equivalent to the category of characters $\psi$ of $G(\Fq)$.
We have also just seen that if the character of $\Lang^*\qcs{E}$ is $\psi$ then the canonical isomorphism $m^*\psi \iso \psi \boxtimes \psi$ determines the isomorphism $\mu_\qcs{E} : \qcs{E} \to \qcs{E}\boxtimes\qcs{E}$ which is part of the data that determines a quasicharacter sheaf on $G$.
That leads to a perspective on the function-sheaf dictionary common in the literature in which one considers $1$-dimensional local systems $\qcs{E}$ on $G$ for which \emph{there exists} an isomorphism $m^*\qcs{E} \iso \qcs{E} \boxtimes\qcs{E}$; see \cite{kamgarpour:09a}*{Intro}, for example.
As a slight variation, one may also consider $1$-dimensional local systems $\gqcs{L}$ on $\bG$ for which \emph{there exists} an isomorphism $\Frob{G}^*\gqcs{L} \iso \gqcs{L}$ and an isomorphism $\bm^*\gqcs{L} \iso \gqcs{L} \boxtimes\gqcs{L}$.
%One may show that all such $\gqcs{L}$ are determined by extensions of $\bG$ by $\EEx$ which are fixed by Frobenius and that the trace of Frobenius provides a canonical isomorphism from $\Hh^2(\bG,\EEx)^{\Frob{}}$ to $G(\Fq)^*$; {\it cf.\ }
%One may show that all such $\gqcs{L}$ are determined by extensions of $\bG$ by $\EEx$ which are fixed by Frobenius and that the trace of Frobenius provides a canonical isomorphism from $\Hh^2(\bG,\EEx)^{\Frob{}}$ to $G(\Fq)^*$; {\it cf.\ } \cite{kamgarpour:09a}*{Prop 2.15}.

Although the category of quasicharacter sheaves on $G$ does specialize to that common notion when $G$ is connected and of finite type, the common notion is not appropriate if one wishes to extend the dictionary to smooth commutative group schemes over finite fields. 
In particular, as we have already seen in Section~\ref{sec:etale}, for a given quasicharacter sheaf $\qcs{L} = (\gqcs{L},\mu,\phi)$ there may be a different isomorphism $\mu' : \bm^*\gqcs{L} \to \gqcs{L}\boxtimes\gqcs{L}$ for which $\qcs{L}'=(\gqcs{L},\mu',\phi)$ is also a quasicharacter sheaf on $G$, not isomorphic to $\qcs{L}$ in $\QC(G)$.
Indeed, Proposition~\ref{prop:etale-iso} shows that when $G$ is \'etale, $\Hh^2(\bG,\EEx)^{\Weil{}}$ measures the possibilities for $\mu'$.
The common perspective on the function-sheaf dictionary leads only to the part of category of quasicharacter sheaves that is invisible to the trace of Frobenius.

We will see in Section~\ref{sec:main} that, for any smooth commutative group scheme $G$, the part of category of quasicharacter sheaves that is invisible to the trace of Frobenius are measured by $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$.
%\end{remark}

>>>>

By any measure, there are more quasicharacter sheaves for $T$ than quasicharacters of $T(K)$.  
In this regard, we are reminded of the work of David Vogan \cite{vogan:93a}, in which he finds a geometrization of complete Langlands parameters for $p$-adic groups 
and, in the process, is led to study the representations of all the pure rational forms of the $p$-adic group, simultaneously.
A similar phenomenon appears in recent work by Joseph Bernstein in which his geometric Ansatz leads to the study of certain sheaves on the stacky classifying space of the $p$-adic group, resulting in a category which appears to be tied to the representations of all the pure rational forms of the $p$-adic group \cite{bernstein:vogan_conference}.
Indeed, Bernstein has suggested to us that our category of quasicharacters for $T$ may be tied to quasicharacters of all the pure rational forms of $T$.

@@

\begin{proof}
Let $f$, $A$, $V$ and $\psi$ be as above; set $\qcs{E} = (f_! V_H)_\psi$.
Since $A$ is abelian, $\qcs{E}$ is an $\ell$-adic local system on $G$ of rank one.
We must show that comes equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E}\boxtimes\qcs{E}$.
To do this we use \'etale descent to see that pullback along $f$ gives an equivalence between local systems on $G$ and $A$-equivariant local systems on $H$; {\it cf.\ } \cite{bernstein-luntz:equivariant_sheaves}*{Prop 8.1.1}. 
In particular, $f^*\qcs{E}$ is the $A$-equivariant constant sheaf $V$ on $H$ with character $\psi$.
Since $f$ is a morphism of group schemes, the functor $f^*$ defines $\mu_\qcs{E} : m^*\qcs{E} \to \qcs{E}\boxtimes\qcs{E}$
from the isomorphism $m^*\psi \iso \psi \boxtimes\psi$ determined by $V\to V\otimes V$.
\end{proof}

@@

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object of $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function-sheaf dictionary for connected commutative algebraic groups over finite fields, as in \cite{deligne:SGA4.5}*{Sommes trig.} or \cite{laumon:87a}*{1.1.3}, gives us all we need, as we briefly recall. %\todo{Fix citelist} %\citelist{\cite{deligne:SGA4.5}*{Sommes trig.},\cite{laumon:87a}*{1.1.3}}
The key observation is that, by \'etale descent, pullback by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ (this is a particular case of the argument in the proof of Proposition~\ref{prop:finite}) and that, for every $\qcs{E}$ in $\bQC(G)$, the trace of $\Lang^*\qcs{E}$ as a representation of $G(\Fq)$, is $\trFrob{\qcs{E}}^{-1}$.

Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_! V)_\psi$, the $\psi$-isotypic component of the push-forward of the constant sheaf $V$ on $G$ by $\Lang$, equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$. 
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
\end{proof}

%\begin{remark}
We have just seen that, for a connected commutative algebraic group $G$ over $\Fq$, the category of quasicharacter sheaves on $G$ is equivalent to the category of one-dimensional representations $V$ of $G(\Fq)$ equipped with an isomorphism $V\iso \EE$, and therefore equivalent to the category of characters $\psi$ of $G(\Fq)$.
We have also just seen that if the character of $\Lang^*\qcs{E}$ is $\psi$ then the canonical isomorphism $m^*\psi \iso \psi \boxtimes \psi$ determines the isomorphism $\mu_\qcs{E} : \qcs{E} \to \qcs{E}\boxtimes\qcs{E}$ which is part of the data that determines a quasicharacter sheaf on $G$.
That leads to a perspective on the function-sheaf dictionary common in the literature in which one considers one-dimensional local systems $\qcs{E}$ on $G$ for which \emph{there exists} an isomorphism $m^*\qcs{E} \iso \qcs{E} \boxtimes\qcs{E}$; see \cite{kamgarpour:09a}*{Intro}, for example.
As a slight variation, one may also consider one-dimensional local systems $\gqcs{L}$ on $\bG$ for which \emph{there exists} an isomorphism $\Frob{G}^*\gqcs{L} \iso \gqcs{L}$ and an isomorphism $\bm^*\gqcs{L} \iso \gqcs{L} \boxtimes\gqcs{L}$.
%One may show that all such $\gqcs{L}$ are determined by extensions of $\bG$ by $\EEx$ which are fixed by Frobenius and that the trace of Frobenius provides a canonical isomorphism from $\Hh^2(\bG,\EEx)^{\Frob{}}$ to $G(\Fq)^*$; {\it cf.\ }
%One may show that all such $\gqcs{L}$ are determined by extensions of $\bG$ by $\EEx$ which are fixed by Frobenius and that the trace of Frobenius provides a canonical isomorphism from $\Hh^2(\bG,\EEx)^{\Frob{}}$ to $G(\Fq)^*$; {\it cf.\ } \cite{kamgarpour:09a}*{Prop 2.15}.

Although the category of quasicharacter sheaves on $G$ does specialize to that common notion when $G$ is connected and of finite type, the common notion is not appropriate if one wishes to extend the dictionary to smooth commutative group schemes. 
In particular, as we have already seen in Section~\ref{sec:etale}, for a given quasicharacter sheaf $\qcs{L} = (\gqcs{L},\mu,\phi)$ there may be a different isomorphism $\mu' : \bm^*\gqcs{L} \to \gqcs{L}\boxtimes\gqcs{L}$ for which $\qcs{L}'=(\gqcs{L},\mu',\phi)$ is also a quasicharacter sheaf on $G$, not isomorphic to $\qcs{L}$ in $\QC(G)$.
Indeed, Proposition~\ref{prop:etale-iso} shows that when $G$ is \'etale, $\Hh^2(\bG,\EEx)^{\Weil{}}$ measures the possibilities for $\mu'$.
On \'etale group schemes, the common perspective on the function-sheaf dictionary leads only to the part of category of quasicharacter sheaves that is invisible to the trace of Frobenius.

We will see in Section~\ref{sec:main} that, for any smooth commutative group scheme $G$, the part of the category of quasicharacter sheaves that is invisible to the trace of Frobenius is measured by $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$.
%\end{remark}
